Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-02-06 paper.qmd
Updated long edge removal section

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two \\gD{} projections from $7\\text{-}D$, for the two non-linear clusters data (grey points) and true model (blue lines) are shown. The top cluster in (a) shows uniformly distributed data along a C-shape, while in the bottom cluster, the data is more concentrated in one corner of the curvy shape. Video of the langevitour animations are available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "We start with a tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 7$), from which the \\gD{} wireframe is constructed (b) and then lifted to $7\\text{-}D$. A \\gD{} projection of the model fit from a tour through the $7\\text{-}D$ space is presented in c. Additionally, a \\gD{} projection of the fitted and true models (d) from the same tour demonstrates that the model accurately captures the non-linear structure of the data, although it does not fully represent the width of that structure. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-parameter) choice. The default is used. Video of the langevitour animation is available at <>."
#| fig-width: 10
#| fig-height: 10

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $7\\text{-}D$ model error in \\gD{} layout. Color indicates square root of total error, dark colour indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the non-linear cluster with dense corner."
#| out-width: 60%
#| fig-pos: H

@@

Edges define the neighbourhood structure, in order to provide a smooth \gD{} representation of the fitted model. @fig-two-curvy-true-proj shows a wire frame of the true model that was used to generate the two C-shaped clusters example data. The ideal is that the representation of the fitted model, at least for this example where we know the true model, should look similar to this.

@@

Another usability of our approach is how NLDR methods behave with different density of the data structure. The tSNE layout of simulated $7\text{-}D$ C-shaped data with different densities are used. The high \pD{} model errors are scattered to a corner of the \gD{} layout of dense C-shaped structure. In the dense C-shaped structure, the high model errors occur because there are fewer data points positioned within the bins at the sparse end of the structure.

>>>>

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two \\gD{} projections from $4\\text{-}D$, for the two non-linear clusters data (grey points) and true model (blue lines) are shown. The top cluster in (a) shows uniformly distributed data along a C-shape, while in the bottom cluster, the data is more concentrated in one corner of the curvy shape. Video of the langevitour animations are available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "We start with a tSNE layout (a) of non-linear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (b) and then lifted to $4\\text{-}D$. A \\gD{} projection of the model fit from a tour through the $4\\text{-}D$ space is presented in c. Additionally, a \\gD{} projection of the fitted and true models (d) from the same tour demonstrates that the model accurately captures the non-linear structure of the data, although it does not fully represent the width of that structure. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-parameter) choice. The default is used. Video of the langevitour animation is available at <>."
#| fig-width: 10
#| fig-height: 10

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates square root of total error, dark colour indicating high error and light indicates low error. Most large errors are distributed near the sparse end of the non-linear cluster with dense corner."
#| out-width: 60%
#| fig-pos: H

@@

Edges define the neighbourhood structure, in order to provide a smooth \gD{} representation of the fitted model. @fig-two-curvy-true-proj shows a wire frame of the true model that was used to generate the non-linear clusters example data. The ideal is that the representation of the fitted model, at least for this example where we know the true model, should look similar to this.

@@

Another usability of our approach is how NLDR methods behave with different density of the data structure. The tSNE layout of simulated $4\text{-}D$ C-shaped data with different densities are used. The high \pD{} model errors are scattered to a corner of the \gD{} layout of dense C-shaped structure. In the dense C-shaped structure, the high model errors occur because there are fewer data points positioned within the bins at the sparse end of the structure.
